{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-05-03 paper.tex
features of fk tables

Given the evolution operator, we may replace the (general-scale) PDFs
used in the subprocess parton density Eq.~(\ref{eq:APPLsubproc}) with
their equivalent expressions c.f Eq.~(\ref{eq:fastPDFfinal_recalled})
like so:
\begin{equation}\label{eq:FK1}
\begin{array}{rcl}
F^{(s)}_{\alpha\beta,\tau} &=&  \displaystyle \sum_{i,j} \sum_{k,l}
                               \sum_{\delta,\gamma} C^{(s)}_{ij}
                               \left[  A^\tau_{\alpha\delta ik}\;
                               f_k(x_\delta,Q^2_0) A^\tau_{\beta\gamma
                               jl}\; f_l(x_\gamma,Q^2_0) \right]\;\;\;
  \\
\\
&=& \displaystyle \sum_{k,l}\sum_{\delta,\gamma}
\widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,,
\end{array}
\end{equation}
where the object:
\begin{equation}
  \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta} =
  \sum_{i,j} C^{(s)}_{ij} A^\tau_{\alpha\delta ik}
  A^\tau_{\beta\gamma jl}
\end{equation}
combines the operations of subprocess density construction and PDF
evolution. Going further it is possible to then use the expression for
subprocess parton densities in Eq.~(\ref{eq:FK1}) in the full
cross-section calculation:
\begin{equation}
\sigma_{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma}\sum_p
\sum_{s} \sum_{\alpha,\beta}
\sum_{\tau} 
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,.
\end{equation}
Performing some further contractions it is possible to obtain an
extremely compact expression for the calculation of the cross section
in question, in terms of only the initial-scale PDFs and summing only
over the initial scale interpolating $x$-grid and the incoming parton
flavors:
\begin{equation}\label{eq:FK}
  \sigma _{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma} 
  \widetilde{W}_{kl,\delta\gamma} \,f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0),
\end{equation}
where the object:
\begin{equation}\label{eq:FKTable}
  \widetilde{W}_{kl,\delta\gamma} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta} \sum_{\tau}
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)  W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
\end{equation}
is referred to here as an {\tt FKtable}, and combines the information
stored in {\tt APPLgrid}-style interpolated weight grids with
analogously interpolated DGLAP evolution operators. Such a combination
enables for a maximally computationally-efficient expression for the
calculation of observables at hadron colliders.

\subsection{Properties of {\tt FK} tables}
Some discussion here of the expected performance
differences. Including precomputed sums over scale, perturbative
order, no need to perform PDF evolution on the fly, smaller initial
scale PDF basis, simple convolution structure that can be accelerated
with SIMD/OpenMP.  Explain reduced flexibility w.r.t $\alpha_S$
variations and scales, discuss increased storage space requirement but
also reduced memory footprint.  Might want to mention possibility for
generating 'fast' grids as the interpolation accuracy is decoupled.

\section{Performance benchmarks}
\label{sec:benchmark}

The {\tt APFELgrid} plugin generates {\tt FK} tables of the form
expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt
  APPLgrid} files. In this section we shall examine the performance
differences between the two expressions for cross-section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt FK}).
%
To provide a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production
measurements~\cite{Aaij:2012mda,Aaij:2012vn,Chatrchyan:2013mza,Chatrchyan:2013uja,Chatrchyan:2012xt,Aad:2013iua,Aad:2011fp,Aad:2011dm,Aaltonen:2010zza},
$t\bar{t}$ total
cross-sections~\cite{ATLAS:2012aa,ATLAS:2011xha,TheATLAScollaboration:2013dja,Chatrchyan:2013faa,Chatrchyan:2012bra,Chatrchyan:2012ria},
double-differential Drell-Yan
cross-sections~\cite{Chatrchyan:2013tia,CMS:2014jea} and inclusive jet
data~\cite{Chatrchyan:2012bja,Aad:2011fc,Aad:2013lpa,Abazov:2007jy}. Predictions
are performed over a wide range of kinematics, for a total of 52 source
{\tt APPLgrid} files corresponding to the majority of available LHC and Tevatron datasets of applicability to PDF determination.
While the source {\tt APPLgrid} files have varying grid densities in $x$ and $Q^2$, for the purposes of comparison the corresponding {\tt FK} tables are produced
consistently with 30 points in $x$, and at an initial scale below the charm threshold, therefore with seven active partonic species.

@@

In this work we have presented {\tt APFELgrid}, an efficient
convolution interface to hadronic observables for fits of parton
densities. We have demonstrated that employing the so-called {\tt
  FastKernel} method, it is possible to convert an {\tt APPLgrid}
weight table into a derived format, referred to as {\tt FKtable},
which includes PDF and $\alpha_s$ evolution as provided by {\tt
  APFEL}.  Such a procedure allows one to obtain a very
computationally-efficient expression for the calculation of hadronic
cross sections, in terms of only the initial-scale PDFs and summing
only over the initial scale interpolating grid and the incoming parton
flavors. In addition, the simple structure of the resulting
convolution makes {\tt FKtables} particularly suitable to make
efficient use of acceleration and parallelization techniques like {\tt
  SIMD} and {\tt OpenMD}.

We have shown that the numerical convolution of an {\tt APFELgrid}
tables is typically much faster than the corresponding {\tt
  APPLgrid}. In particular, the {\tt FastKernel} method provides a
great perfomance improvement with respect to {\tt APPLgrid} which is
further enhanced by the employment of any of the acceleration
techniques available in {\tt APFELgrid}. Finally, while the size of an
{\tt FKtable} is often larger than the corresponding {\tt APFELgrid}
file, we have shown that the resident set size memory required by {\tt
  APFELgrid} is typically smaller than {\tt APPLgrid} by at least
75\%.

The subtantial speed improvement with respect to {\tt APPLgrid} in
association with the reduction of the resident set size memory, makes
the {\tt APFELgrid} code a valuable tool for modern PDF fits which
usually incluse thousands of hadronic data points and that thus
heavily rely on fast interfaces. We stress that {\tt APFELgrid} is not
a replacement for {\tt APPLgrid}, on which it depends, but it is
rather a specialized tool conceived to ease the inclusion of hadronic
data into PDF fits. In fact, the price of a better performance is the
fact that, by themselves, {\tt FKtables} cannot perform any parameter
variation other than that of PDFs. For example, variations of
$\alpha_s$ can be performed only by re-computing the {\tt FKtables}.

>>>>

Given the evolution operator, we may replace the (general-scale) PDFs
used in the subprocess parton density Eq.~(\ref{eq:APPLsubproc}) with
their equivalent expressions c.f Eq.~(\ref{eq:fastPDFfinal_recalled})
like so:
\begin{equation}\label{eq:FK1}
\begin{array}{rcl}
F^{(s)}_{\alpha\beta,\tau} &=&  \displaystyle \sum_{i,j} \sum_{k,l}
                               \sum_{\delta,\gamma} C^{(s)}_{ij}
                               \left[  A^\tau_{\alpha\delta ik}\;
                               f_k(x_\delta,Q^2_0) A^\tau_{\beta\gamma
                               jl}\; f_l(x_\gamma,Q^2_0) \right]\;\;\;
  \\
\\
&=& \displaystyle \sum_{k,l}\sum_{\delta,\gamma}
\widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,,
\end{array}
\end{equation}
where the object:
\begin{equation}
  \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta} =
  \sum_{i,j} C^{(s)}_{ij} A^\tau_{\alpha\delta ik}
  A^\tau_{\beta\gamma jl}
\end{equation}
combines the operations of subprocess density construction and PDF
evolution. Going further it is possible to then use the expression for
subprocess parton densities in Eq.~(\ref{eq:FK1}) in the full
cross-section calculation:
\begin{equation}
\sigma_{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma}\sum_p
\sum_{s} \sum_{\alpha,\beta}
\sum_{\tau} 
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,.
\end{equation}
Performing some further contractions it is possible to obtain an
extremely compact expression for the calculation of the cross section
in question, in terms of only the initial-scale PDFs and summing only
over the initial scale interpolating $x$-grid and the incoming parton
flavors:
\begin{equation}\label{eq:FK}
  \sigma _{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma} 
  \widetilde{W}_{kl,\delta\gamma} \,f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0),
\end{equation}
where the object:
\begin{equation}\label{eq:FKTable}
  \widetilde{W}_{kl,\delta\gamma} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta} \sum_{\tau}
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)  W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
\end{equation}
is referred to here as an {\tt FK} table, and combines the information
stored in {\tt APPLgrid}-style interpolated weight grids with
analogously interpolated DGLAP evolution operators. This combination
enables for a maximally computationally-efficient expression for the
calculation of observables at hadron colliders under PDF variation.

\subsection{Features and limitations of {\tt FK} tables}
The {\tt FK} product of Eqn.~\ref{eq:FK} differs with respect to the product in Eqn.~\ref{eq:applconv} in several notable ways. Firstly 
the typical {\tt APPLgrid} or {\tt FastNLO} products require PDFs at a general scale $Q^2$, which implies performing PDF evolution e.g Eqn.~\ref{eq:fastPDFfinal_recalled} for every variation of the PDFs at the input scale during the fit. However in the {\tt FK} convolution this evolution
is pre-cached at the stage of {\tt FK} table generation, requiring only initial-scale PDFs as an input at fitting time. This pre-caching of the evolution
also removes the need for performing the sums over hard scale and perturbative order during the fit, further reducing the number of operations required.
The {\tt FK} convolution also benefits from the typically reduced number of active partonic modes at the fitting scale, with the sum over flavours in Eqn.~\ref{eq:FK} being limited to those directly used in the fit. Having reduced the calculation to such a simple form, it is also possible to straightforwardly
apply standard and widely-available computational tools such as multi-threading through e.g OpenMP or Single Instruction Multiple Data (SIMD) operations e.g SSE or AVX to further reduce computational cost.

While these features provide significant performance enhancements, the {\tt FK } table format is not suitable as a complete replacement for tools
such as {\tt APPLgrid}. The pre-computation of the PDF evolution necessarily means that all theory parameters such as perturbative order, strong coupling and factorization/renormalization scales are inextricably embedded in each {\tt FK} table. In order to perform PDF fits including variations of these parameters, multiple {\tt FK } tables must be computed, each with different theory settings. While performing such a re-calculation directly from Monte Carlo codes
would be exceptionally time consuming, the data representation in {\tt APPLgrid} files allows for an efficient combination with varying theory parameters.

%\section{The {\tt APFELgrid} package}
% **** Not sure if this sort of section would be neccesary ***
%The {\tt APFELgrid} plugin generates {\tt FK} tables of the form
%expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt
%  APPLgrid} files. It is supplied as a set of {\tt C++} routines designed to
%  supplement the PDF evolution library {\tt APFEL} with {\tt FK} table generation
%  capabilities. The plugin has a very light interface for performing the {\tt FK} combination,
%  and relies on the native {\tt APFEL} API for setting the parameters of the PDF evolution,
%  including the structure of the initial scale $x$ grid.
%*** Would now discuss e.g SIMD implementation etc in advance of looking at the results in next section ***
\section{Performance benchmarks}
\label{sec:benchmark}

In this section we shall examine the performance
differences between the two expressions for cross-section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt FK}).
%
To provide a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production
measurements~\cite{Aaij:2012mda,Aaij:2012vn,Chatrchyan:2013mza,Chatrchyan:2013uja,Chatrchyan:2012xt,Aad:2013iua,Aad:2011fp,Aad:2011dm,Aaltonen:2010zza},
$t\bar{t}$ total
cross-sections~\cite{ATLAS:2012aa,ATLAS:2011xha,TheATLAScollaboration:2013dja,Chatrchyan:2013faa,Chatrchyan:2012bra,Chatrchyan:2012ria},
double-differential Drell-Yan
cross-sections~\cite{Chatrchyan:2013tia,CMS:2014jea} and inclusive jet
data~\cite{Chatrchyan:2012bja,Aad:2011fc,Aad:2013lpa,Abazov:2007jy}. Predictions
are performed over a wide range of kinematics, for a total of 52 source
{\tt APPLgrid} files corresponding to the majority of available LHC and Tevatron datasets of applicability to PDF determination.
While the source {\tt APPLgrid} files have varying grid densities in $x$ and $Q^2$, for the purposes of comparison the corresponding {\tt FK} tables are produced
consistently with 30 points in $x$, and at an initial scale below the charm threshold, therefore with seven active partonic species.

@@

In this work we have presented {\tt APFELgrid}, an efficient
convolution interface to hadronic observables for fits of parton
densities. We have demonstrated that employing the so-called {\tt
  FastKernel} method, it is possible to convert an {\tt APPLgrid}
weight table into a derived format, referred to as {\tt FKtable},
which includes PDF and $\alpha_s$ evolution as provided by {\tt
  APFEL}.  Such a procedure allows one to obtain a very
computationally-efficient expression for the calculation of hadronic
cross sections, in terms of only the initial-scale PDFs and summing
only over the initial scale interpolating grid and the incoming parton
flavors. In addition, the simple structure of the resulting
convolution makes {\tt FKtables} particularly suitable to make
efficient use of acceleration and parallelisation techniques like {\tt
  SIMD} and {\tt OpenMP}.

We have shown that the numerical convolution of an {\tt APFELgrid}
tables is typically much faster than the corresponding {\tt
  APPLgrid}. In particular, the {\tt FastKernel} method provides a
great performance improvement with respect to {\tt APPLgrid} which is
further enhanced by the employment of any of the acceleration
techniques available in {\tt APFELgrid}. Finally, while the size of an
{\tt FKtable} is often larger than the corresponding {\tt APFELgrid}
file, we have shown that the resident set size memory required by {\tt
  APFELgrid} is typically smaller than {\tt APPLgrid} by at least
75\%.

The substantial speed improvement with respect to {\tt APPLgrid} in
association with the reduction of the resident set size memory, makes
the {\tt APFELgrid} code a valuable tool for modern PDF fits which
usually incluse thousands of hadronic data points and that thus
heavily rely on fast interfaces. We stress that {\tt APFELgrid} is not
a replacement for {\tt APPLgrid}, on which it depends, but it is
rather a specialised tool conceived to ease the inclusion of hadronic
data into PDF fits. In fact, the price of a better performance is the
fact that, by themselves, {\tt FKtables} cannot perform any parameter
variation other than that of PDFs. For example, variations of
$\alpha_s$ can be performed only by re-computing the {\tt FKtables}.
